---
title: "IS607 - Project 2"
author: "Mauricio Alarcon, Jamey Etherton"
date: "March 3, 2015"
output: html_document
---

* * *

##<a name="toc"></a>Table of Contents
* [Introduction](#introduction)
* [Methodology](#methodology)
* Exploratory Analysis
    * [Descriptive Statistics](#descriptive_statistics)
    * [y=f(x)](#y_fx)
        * [Group I](#group_i)
        * [Group II](#group_ii)
        * [Group III](#group_iii)
        * [Group IV](#group_iv)
* [Data Modeling](#data_modeling)
    * [Group I](#group_i_m)
    * [Group II](#group_ii_m)
    * [Group III](#group_iii_m)
    * [Group IV](#group_iv_m)
* [Conclusion](#conclusion)

* * *

```{r, echo=FALSE,results = 'hide', warning=FALSE, message=FALSE}
library("ggplot2")
library("boot")
library("pastecs")
library("reshape")
library("lattice")
library("Hmisc")
library("reshape2")
library("RCurl")

# get files from CSV in github
project2_data_csv <- getURL("https://raw.githubusercontent.com/rmalarc/is607-project2/master/project2_data.csv")
project2_data <- read.csv(text=project2_data_csv,head=TRUE,sep=",",as.is=TRUE)

```


##<a name="introduction"></a>Introduction

As part of the IS606 class, we're provided with a PDF file containing a data set. The data set appears to contain two variables (X,Y), and the respective values across four groups (I,II,III,IV)

We're given this data for a reason: to Analyze it!.

[[Back to Top]](#toc)

* * *

##<a name="methodology"></a>Methodology

The methodology of this analysis consists of:

* Capture provided data in CSV format with the following columns: group, x, y
* Load the data into R
* Conduct exploratory analysis including: 
    * Obtain descriptive statistics by group
    * Produce basic y~x charts including linear fitting curve
    * Result interpretation
* Data Modeling

[[Back to Top]](#toc)

* * *
##Exploratory Analysis

###<a name="descriptive_statistics"></a>Descriptive Statistics

#### X Variable
```{r, echo=FALSE}

qplot(group,x, data=project2_data, geom=c("boxplot"),
      fill=group,
      main="Distribution of X Values by Group",
      xlab= "Group",
      ylab="X")+
  scale_fill_discrete(name="Group")+ guides(fill=FALSE)

options(digits=2)
aggregate(project2_data$x,list(Group=project2_data$group),FUN=summary)

# Show X-Variable values and frequency of each
dcast(project2_data, group ~ x, value.var="x",fun.aggregate=length)

```

From the summary and plot above we can see how the values of x appear to be identical in groups I, II and III, with X being an integer uniformly distributed with values spanning from 4 to 14.

Group IV however, has only two numerical values: Y=10 (n=10) and Y=19 (n=1).

[[Back to Top]](#toc)

#### Y Variable
```{r, echo=FALSE}

qplot(group,y, data=project2_data, geom=c("boxplot"),
      fill=group,
      main="Distribution of Y Values by Group",
      xlab= "Group",
      ylab="Y")+
  scale_fill_discrete(name="Group")+ guides(fill=FALSE)

aggregate(project2_data$y,list(Group=project2_data$group),FUN=summary)

```

Aside from the fact that the mean of Y across all groups is 7.5, the distribution of its values is different across groups.

[[Back to Top]](#toc)

###<a name="y_fx"></a>y=f(x)

Given the fact that there are differences in the basic distribution of X,Y values, we break the analysis of these combinations for each of the groups (I,II,III,IV).

For this part of the analysis we plot the (x,y) value pairs alongside a linear approximation.

####<a name="group_i"></a>Group I

```{r, echo=FALSE}
ggplotRegression <- function (fit) {

require(ggplot2)

ggplot(fit$model, aes_string(x = names(fit$model)[2], y = names(fit$model)[1])) + 
  geom_point() +
  stat_smooth(method = "lm", col = "red") +
  labs(title = paste("y=",signif(fit$coef[[2]], 2),
                     "x +",signif(fit$coef[[1]],2 ),
                     ", Adj R2=",signif(summary(fit)$adj.r.squared, 5),
                     ", p=",signif(summary(fit)$coef[2,4], 5)))
}


fit1 <- lm(y ~ x, data = project2_data[project2_data$group=="I",])
ggplotRegression(fit1)


#qplot(x,y, data=project2_data[project2_data$group=="I",],
#      color=group, shape=group,
#      geom=c("point", "smooth"),
#      method="lm", formula=y~x,
#      xlab="X", ylab="Y",
#      main="Regression of Values X~Y for Group I")
```

From the above chart we can see that the data in group I does follow a linear trend, with acceptable R2 and p-values.

[[Back to Top]](#toc)

####<a name="group_ii"></a>Group II

```{r, echo=FALSE}
fit1 <- lm(y ~ x, data = project2_data[project2_data$group=="II",])
ggplotRegression(fit1)
```

Here we see that group II's regression output are identical to those of group I. However, it is evident that a quadratic approximation may be a better fit.

[[Back to Top]](#toc)

####<a name="group_iii"></a>Group III

```{r, echo=FALSE}

fit1 <- lm(y ~ x, data = project2_data[project2_data$group=="III",])
ggplotRegression(fit1)

```

In this case we can see the presence of an outlier (13,12.7). This value clearly skews the curve. If removed, a much better fit can be obtained.

[[Back to Top]](#toc)

####<a name="group_iv"></a>Group IV

```{r, echo=FALSE}

fit1 <- lm(y ~ x, data = project2_data[project2_data$group=="IV",])
ggplotRegression(fit1)


```

This series appears to include multiple Y observations for X=8. The average of these multiple Y values can be used instead.

However, it does not appear that the resulting function would be much different than y=0.5x + 3.

[[Back to Top]](#toc)

* * *
##<a name="data_modeling"></a>Data Modeling

[[Back to Top]](#toc)

* * *

##<a name="conclusion"></a>Conclusion

[[Back to Top]](#toc)
